\documentclass[12pt, a4paper]{report}
\input{../preamble.tex}

\title{CSCA67H3 - Discrete Mathematics \\ Week 1 Tutorial Questions}
\author{Gosha Polstvin}
\date{Year 1 — Fall 2025}

\begin{document}
\maketitle

\chapter*{Solutions}

\section{Exercises 1}

\emph{Either both Ralph and Ed are tall, or both of them are handsome.}

\textbf{Solution:}
\[
    (R_t \land E_t) \lor (R_h \land E_h)
. \]\\

\emph{Both Ralph and Ed are either tall or handsome.}

\textbf{Solution:}
\[
    (R_t \lor R_h) \land (E_t \lor E_h)
.\] \\

\emph{Both Ralph and Ed are neither tall nor handsome.}

\textbf{Solution:}
\[
    (\neg R_t \land \neg R_h) \land (\neg E_t \land \neg E_h)
.\] \\

\emph{Neither Ralph nor Ed is both tall and handsome.}

\textbf{Solution:} This one was a bit tricky. The natural reading is that no one has both properties: 
\[
    \neg(R_t \land R_h) \land \neg(E_t \land E_h)
.\] \\

\emph{Alternative reading:} It could also be interpreted as “it's false that
both of them simultaneously have both properties." Basically, not all of the
following conditions can be true at the same time: 
\[
    \neg(R_t \land R_h \land E_t \land E_h)
.\] \\

\emph{Either both Ralph and Ed are tall, or neither of them are.}

\textbf{Solution:}
\[
    (R_t \land E_t) \lor (\neg R_t \land \neg E_t)
.\] \\

\emph{Either Ralph is tall and Ed is handsome, or vice versa.}

\textbf{Solution:}
\[
    (R_t \land E_h) \lor (E_t \land R_h)
\]

\section{Exercises 2}

Let \(T\) denote “taxes will go up” and \(D\) denote “the deficit will go up.”  

\[
T \lor D
.\] \\

\emph{Taxes will go up or the deficit will go up.}

\[
(T \lor D) \land \neg(T \land D)
.\] \\

\emph{Either taxes will go up or the deficit will, but not both.}

\[
(T \land \neg D) \lor (D \land \neg T)
.\] \\

\emph{Either taxes will go up but not the deficit, or the deficit will go up but not taxes.}

\end{document}

